Matrix Transposition and Symmetry: a case study on Superscalar Architectures\\

### 2024-11-27 latex/paper.tex
report: completed section 3

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{blindtext}  % generate random text
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

@@

Computing the transpose of a matrix is a well discussed computational problem. Early
work focuses on in-situ transporision, that is transposing a matrix without
relying on additional space. Implementations were often based on the cyclic
structure of the transposition mapping: values of a matrix of size $N$ can be stored
in a single contiguous array or size $N$*$N$ and indexed with $row*\#rows+ column$.
Remarcable contributions include Boothroyd's ACM Algorithm 302 (1966) using cyclic mapping \cite{b13}, Laflin and Brefner ACM Algorithm 380
adding fixed points arithmetic \cite{b14} and
a later revision of the two by Esko G. Cate and David W. Twigg in ACM Algorithm 513 \cite{b15} .
SIMD implementations for specific architectures are also widely adopted
in performance critical applications like cryptographic libraries using instructions like Intel's vpunpckldq and vpunpckhdq \cite{b16} reporting improvements up to 70\% \cite{b17}.
However, matrix operations are not efficient as vector registers access data only linearly, motivating research on new vector register file (VFS) designs such as diagonal registers \cite{b23}.
Another area of research includes out-of-core matrix transposition, that
is computing the transpose of a matrix that exceeds the size of the available in-core memory.
This problem focuses on optimizing the number of I/O operations and file layout since memory-memory data transfer is notably slower than math operations. Work in out-of-core transposition include W. 0. Alltop (1975) \cite{b18}, Suh \& Prasanna (2002) \cite{b19} and Krishnamoorth et al. (2004) \cite{b20}.
Recent work explores parallelization and the use of GPUs to offload algebra operations. Examples are I-Jui Sung et al \cite{b21} and Mark Harris \cite{b22}.

\textit{A square matrix $A$ is called symmetric if it equals its transpose} \cite{b12}.

While symmetric matrces are actively used in various areas of research as their
properties allow for several optimizations in various problems, an in depth case study of symmetry
checking on modern architectures is yet to be published.

@@

- Elaborate on the unique contributions of your project \\
- Describe the methodology proposed, including algorithms (pseudo-code),
data structures, and parallelization techniques. \\
- Discuss the challenges faced and how they were addressed \\

@@

Quiesti qui bellissimi e stra fighi
- uno (pseudocodice)
- due (pseudocodice)
- tre (pseudocodice)
- ...
Ho usato implicit ed esplicit parallelism con vectorization e openmp
Voglio fare questo per comparare quest'altro cos' pel meme
Benchmarkare è difficle perchè... quindi ho fatto ciò... (aplia qua)

@@

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{blindtext}  % generate random text
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}
% \usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{algpseudocode}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\iffalse

@@

Computing the transpose of a matrix is a well discussed computational problem. Early
work focuses on in-situ transporision, that is transposing a matrix without
relying on additional space. Implementations were often based on the cyclic
structure of the transposition mapping: values of a matrix of size $N$ can be stored
in a single contiguous array or size $N$*$N$ and indexed with $row*\#rows+ column$.
Remarcable contributions include Boothroyd's ACM Algorithm 302 (1966) using cyclic mapping \cite{b13}, Laflin and Brefner ACM Algorithm 380
adding fixed points arithmetic \cite{b14} and
a later revision of the two by Esko G. Cate and David W. Twigg in ACM Algorithm 513 \cite{b15} .
SIMD implementations for specific architectures are also widely adopted
in performance critical applications like cryptographic libraries, utilizing instructions like Intel's vpunpckldq and vpunpckhdq \cite{b16} and reporting improvements up to 70\% \cite{b17}.
However, matrix operations are not efficient as vector registers access data only linearly, motivating research on new vector register file (VFS) designs such as diagonal registers \cite{b23}.
Another area of research includes out-of-core matrix transposition, that
is computing the transpose of a matrix that exceeds the size of the available in-core memory.
This problem focuses on optimizing the number of I/O operations and file layout since memory-memory data transfer is notably slower than math operations. Work in out-of-core transposition include W. 0. Alltop (1975) \cite{b18}, Suh \& Prasanna (2002) \cite{b19} and Krishnamoorth et al. (2004) \cite{b20}.
Recent work explores parallelization and the use of GPUs to offload algebra operations. Examples are I-Jui Sung et al \cite{b21} and Mark Harris \cite{b22}.

\iffalse
\textit{A square matrix $A$ is called symmetric if it equals its transpose} \cite{b12}.
\fi 
While symmetric matrces are actively used in various areas of research as their
properties allow for several optimizations in various problems, an in depth case study of symmetry
checking on modern architectures is yet to be published.

@@

\iffalse % pseudocode
    \begin{algorithmic}[1]
      \Procedure{MyProcedure}{}
      % use \state and \textit{} for variables
      % use \text for text
      % use \gets for assignments
      % use \If \EndIf \Return \Else
      \EndProcedure
    \end{algorithmic}
\fi

In this report different implementations of matTranspose
and checkSymm are benchmarked, comparing implicit and explicit parallelism
techniques. In particular, the benchmarks aim to indicate cache and SIMD performance differences,
speedup with multiple threads and the impact of compile flags optimizations. Data will be compared through tables
and visually via graphs such as the roofline model. All benchmarks
will be tested against every matrix size from $2^2$ to $2^{12}$, we will assume the size value is a multiple of 4 for ease of implementation.  \\
The algorithms for \textit{matTranspose} will now be explained.
Each algorithm takes a $N \times N$ input square matrix \textit{input}, a matrix of the same shape \textit{ouput}, and the size of the two matrices \textit{size}.
The analyzed algorithms are the following:
\begin{itemize}
\item \textit{matTranspose}: naive implementation of matTranspose, iterating over values of the input matrix and storing them in the output matrix with indexes of rows and columns swapped. See Algorithm~\ref{matTranspose} for the pseudocode.

\begin{algorithm}
	\caption{matTranspose}\label{matTranspose}
	\begin{algorithmic}[1]
		\For{\textit{r} \textbf{ in } \textit{0..size-1}}
	    \For{\textit{c} \textbf{ in } \textit{0..size-1}}
		\State $output[c][r] \gets input[r][c]$
		\EndFor
		\EndFor
	\end{algorithmic}
\end{algorithm}

    \item \textit{matTransposeColumns}: similar to \textit{matTranspose}, but iterating on the input matrix first by columns and then by rows.
  The purpose of this implementation is to test a different access pattern.

\iffalse
\begin{algorithm}
	\caption{matTransposeColumns}\label{matTransposeColumns}
\begin{algorithmic}[1]
	\For{\textit{r} \textbf{ in } \textit{0..size-1}}
	 \For{\textit{c} \textbf{ in } \textit{0..size-1}}
            \State $output[r][c] \gets input[c][r]$
            \EndFor
        \EndFor
\end{algorithmic}
\end{algorithm}
\fi

  \item \textit{matTransposeHalf}: similar to \textit{matTranspose}, but the algorithm
  iterates only on the upper half of the input matrix. For each
  pair $(i,j)$ of indices of rows and columns in the upper half of the input matrix $M$, with output matrix $O$, $O[i][j]=M[j][i]$ and $O[j][i] = M[i][j]$.
\iffalse
\begin{algorithm}
	\caption{matTransposeHalf}\label{matTransposeHalf}
\begin{algorithmic}[1]
	\For{\textit{r} \textbf{ in } \textit{0..size-1}}
           \For{\textit{c} \textbf{ in } \textit{r..size-1}}
            \State $output[r][c] \gets input[c][r]$
            \State $output[c][r] \gets input[r][c]$
            \EndFor
        \EndFor
\end{algorithmic}
\end{algorithm}
\fi

\item \textit{matTransposeCyclic}: implementation of \cite{b13}, storing the matrix as a contiguous one-dimensional array of size $N*N$. See Algorithm~\ref{matTransposeCyclic} for the pseudocode.
  \begin{algorithm}
  	\caption{matTransposeCyclic}\label{matTransposeCyclic}
  \begin{algorithmic}[1]

  	  \For{\textit{n} \textbf{ in } \textit{0..size*size-1}}
        \State $int\ \textit{i} \gets \frac{n}{size}$
        \State $int\ \textit{j} \gets n \mod N$
        \State $T[n] \gets M[size * j + i]$
        \EndFor
          \end{algorithmic}
  \end{algorithm}

\item \textit{matTransposeIntrinsic}: based on \textit{matrix\_transpose4x4} from the linux kernel \cite{b16}, revised to work for out-of-place float matrices on x86 architectures and implemented using intrinsics. \textit{transpose\_4x4\_f32\_intrinsic} transposes a 4 by 4 matrix using vector registers of 128 bits by doing the following steps: load 4 rows of 4 floats from the the input matrix, combine pairs of rows in an alternating pattern (interleaving), shuffle to get the transpose and store them in the output matrix. \textit{matTransposeIntrinsic} iterates over 4x4 blocks in the input matrix, calling \textit{transpose\_4x4\_f32\_intrinsic} and saving the computed transposed block in the output matrix with rows and columns swapped. See Algorithm~\ref{matTransposeIntrinsic} for the pseudocode.
  \begin{algorithm}
	\caption{matTransposeIntrinsic}\label{matTransposeIntrinsic}
	    \begin{algorithmic}[1]
	\iffalse

      \Procedure{transpose\_4x4\_f32\_intrinsic}{Matrix4x4 in, Matrix4x4 out}
    \State \textbf{comment} Load rows  
    \State \textbf{comment} unpack and interleave rows
    \State \textbf{comment} Shuffle to form transposed rows
    \State \textbf{comment} Store transposed rows
      \EndProcedure

        \For{$int\ i = 0, i < size, i += 4$}
        \For{$int\ j = 0, j < size, j += 4$}
        \State $transpose\_4x4\_f32\_intrinsic(input[i][j],$
        \State $                               output[j][i])$
        \EndFor
        \EndFor

    \end{algorithmic}
    \end{algorithm}

\item \textit{matTransposeIntrinsicCyclic}: same as \textit{matTransposeIntrinsic} but the matrix is saved as a one dimensional array similarly to \textit{matTransposeCyclic}
%\item matTranspose vectorization:
%\item matTranspose half vectorization:
\item \textit{matTransposeUnrollingOuter}: similar to \textit{matTranspose} but the outer loop is manually unrolled. See Algorithm~\ref{matTransposeUnrollingOuter} for the pseudocode.

  \begin{algorithm}
    \caption{matTransposeUnrollingOuter}\label{matTransposeUnrollingOuter}
    \begin{algorithmic}[1]
        \For{$int\ c = 0, i < size, i += 4$}
            \For{\textit{r} \textbf{ in } \textit{0..size-1}}
            \State $output[c][r] \gets input[r][c]$
            \State $output[c+1][r] \gets input[r][c+1]$
            \State $output[c+2][r] \gets input[r][c+2]$
            \State $output[c+3][r] \gets input[r][c+3]$
            \EndFor
        \EndFor
    \end{algorithmic}
  \end{algorithm}

\item \textit{matTransposeUnrollingInner}: same as \textit{matTransposeUnrollingOuter} but the inner for loop is manually unrolled

\item \textit{matTransposeHalfUnrollingOuter}: same as \textit{matTransposeHalf} but the outer for loop is manually unrolled

\item \textit{matTransposeHalfUnrollingInner}: same as \textit{matTransposeHalf} but the inner for loop is manually unrolled

\item \textit{matTransposeCyclicUnrolled}: same as \textit{matTransposeCyclic} but the for loop in manually unrolled
\iffalse
  \begin{algorithm}
    \caption{matTranspose cyclic unrolled}\label{matTransposeCyclicUnrolled}
    \begin{algorithmic}[1]
      \Procedure{matTransposeCyclicUnrolled}{$Matrix input, Matrix output, int size$}
        \For{\textit{n} \textbf{ in } \textit{0..size*size-1}}
        \State $T[n] \gets M[size * ((n+1) \mod size ) + (\frac{n}{size})]$
        \State $T[n+1] \gets M[size * ((n+1+1) \mod size ) + (\frac{n+1}{size})]$
        \State $T[n+2] \gets M[size * ((n+2+1) \mod size ) + (\frac{n+2}{size})]$
        \State $T[n+3] \gets M[size * ((n+3+1) \mod size ) + (\frac{n+3}{size})]$
        \EndFor
      \EndProcedure
    \end{algorithmic}
  \end{algorithm}
  \fi
\item \textit{matTransposeOmpX}: this set of algorithms are all derived from \textit{matTranspose} with the outer for loops parallelized using omp with the number of threads \textit{X} being a power of 2 from $2^1$ to $2^6$.

\item \textit{matTransposeOmpXCollapse}: similar to \textit{matTransposeOmpX} but the for loops are parellelized using \textit{collapse(2)}

\item \textit{matTransposeOmp16SchedX}: similar to \textit{matTransposeOmpX} with 16 threads using different schedule algorithms, namely: static, dynamic, guided.
\end{itemize}

The algorithms for \textit{checkSymm} will now be explained. Each algorithm take an $N \times N$ square matrix \textit{input}, a size \textit{size} and return true if the matrix is symmetrix, false otherwise.
The analyzed algorithms are the following:
\begin{itemize}
\item \textit{checkSymm}: naive implementation of checkSymm. It Iterates over the upper half of the matrix and testing if every value is equal to the symmetric value. See Algorithm~\ref{checkSymm} for the pseudocode.

  \begin{algorithm}
    \caption{checkSymm}\label{checkSymm}
    \begin{algorithmic}[1]
        \State $\textbf{bool}\ isSymm \gets true$
        \For{\textit{i} \textbf{ in } \textit{0..size-1}}
            \For{\textit{j} \textbf{ in } \textit{i..size-1}}
            	\If{$input[i][j] != input[j][i]$}
            	\State $isSymm \gets \textbf{false}$
            	\EndIf
        	\EndFor
        \EndFor
        \State \Return $isSymm$
    \end{algorithmic}
  \end{algorithm}

% \item \textit{checkSymColumns}:

  % \item \textit{checkSymVectorization}
\item \textit{checkSymmUnrolledOuter}: similar to \textit{checkSymm} but the outer loop is manually unrolled.

\iffalse
  \begin{algorithm}
    \caption{checkSymmUnrolledOuter}\label{checkSymmUnrolledOuter}
    \begin{algorithmic}[1]
      \Procedure{checkSymmUnrolledOuter}{$Matrix input, int size$}
        \State $isSymm = true$
        \For{\textit{i} \textbf{ in } \textit{0..size-1}}
            \For{\textit{j} \textbf{ in } \textit{i..size-1}}
            \If $input[i][j] != input[j][i]$
            \State $isSymm = false$
            \EndIf
            \If $input[i+1][j] != input[j][i+1]$
            \State $isSymm = false$
                        \EndIf
            \If $input[i+2][j] != input[j][i+2]$
            \State $isSymm = false$
                        \EndIf
            \If $input[i+3][j] != input[j][i+3]$
            \State $isSymm = false$
                        \EndIf
        \EndFor
        \EndFor
        \Return $isSymm$
      \EndProcedure
    \end{algorithmic}
  \end{algorithm}
\fi

\item \textit{checkSymmUnrollingInner}: similar to \textit{checkSymmUnrolledOuter} but the inner loop is manually unrolled instead of the outer loop.
\item \textit{checkSymmUnrollingOuterNoBranch}: similar to \textit{checkSymmUnrollingOuter}, but 4 pairs of values are checked for equality with a single math formula. The formula performs a XOR on 4 pairs of values and their transpose, if any of those are different than 0 then the matrix is not symmetric. See Arlgoritm~\ref{checkSymmUnrollingOuterNoBranch} for the pseudocode.

\begin{algorithm}
	\caption{checkSymmUnrollingOuterNoBranch}\label{checkSymmUnrollingOuterNoBranch}
	\begin{algorithmic}[1]
		\State $\textbf{bool}\ isSymm \gets true$
		\For{$int\ i = 0; i < size; i += 4$}
		\For{$int\ j = i; j < size; j++$}
		\State $isSymm \gets isSymm \mid (((input[i][j] \oplus input[j][i]) \mid (input[i+1][j] \oplus input[j][i+1]) \mid (input[i+2][j] \oplus input[j][i+2]) \mid (input[i+3][j] \oplus input[j][i+3])) != 0)$
		\EndFor
		\EndFor
		\State \Return $isSymm$
	\end{algorithmic}
\end{algorithm}

\item \textit{checkSymmUnrollingInnerNoBranch}: similar to \textit{checkSymmUnrollingOuterNoBranch} but the inner loop is manually unrolled.
\item \textit{checkSymOmpX}: similar to \textit{matTransposeOmpX} but based on \textit{checkSymm}. After each parallel section, a barrier is needed to synchronize the shared boolean output variable.
\item \textit{checkSymOmpXCollapse}: similar to \textit{checkSymmOmpX} bur the two loops are parallelized with collapse(2).
\item \textit{checkSymm16SchedX}: similar to \textit{matTranspose16SchedX} but based on \textit{matTranspose}.
\end{itemize}

@@

@@

The experiments were conducted on a cluster equipped with 64 x86\_64 cpus, more specifically Intel(R) Xeon(R) Gold 5220 CPU @ 2.20GHz.
